---
title: "Report Findings"
author: "Youssef Emam"
date: "`r Sys.Date()`"
output: github_document
always_allow_html: true
---

```{r setup, include=FALSE, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	message = FALSE,
	warning = FALSE
	
)
source("main.R")

```

## Data Overview

------------------------------------------------------------------------

## EDA

### Station Data

Once I loaded the data the next step was to look at it. To do this I completed a basic exploratory data analysis. I created plots that summarized the data to get an idea of the distribution across the different data sets.

Firstly I summarized each station by the amount of traffic that it gets. In this case I defined traffic as how many times a trip starts or stops at a station. The plot can be seen below:

```{r traffic, echo=FALSE, include=TRUE}
#create a figure that summarizes station data
ggplot(data = stations_summary, mapping = aes(x = reorder(name, traffic), y = traffic, fill = traffic)) +
  geom_col() +
  coord_flip()+
  theme(axis.text.y = element_text(size = 4, vjust = 0.5),
        plot.title = element_text(hjust = 0.5)) +
  labs(y = "Traffic", x = "Station", title = "Station Traffic")
```

It can be seen that there is quite a large variance in how much traffic each station gets.

### Trip Data

Next I summarized the trip data by identifying the average cycling duration over the course of each month. The plot can be seen below:

```{r trips, echo=FALSE, include=TRUE, eval=TRUE}
#plot the summary
# Seems to be outliers, we will deal with this later
ggplot(data = trips_summary, mapping = aes(x = month, y = average_trip, group = 1))+
  geom_line()+
  labs(y = "Average Trip Duration (Seconds)", x = "Month", title = "Average Trip Duration by Month") +
  theme_classic()+
  theme(plot.title = element_text(hjust = 0.5),
        panel.grid.major.y = element_line(color = "gray", linetype = "dashed")) +
  geom_label(aes(label = round(average_trip, 0)),
             vjust = -0.5,
             size = 2)
```

Unexpectedly, the highest mean duration occurs during December. This is likely due to the presence of an outlier which we will investigate later. Aside from this, as expected the usage seems to increase in the summer months and decrease slightly in the winter months.

### Weather data

To summarize the weather data, I identified how the precipitation levels change over the course of the year with respect to each city represented in the data:

```{r weather, echo=FALSE, include=TRUE, eval=TRUE}
#summarise the weather data
#plot a summary
ggplot(data = weather_summary, mapping = aes(x = month, y = precip, color = city, group = city))+
  geom_line() +
  labs(y = "Total Precipitation by Month (In)", x = "Month", title = "Total Precipitation")+
  theme(plot.title = element_text(hjust = 0.5))
```

As expected, the precipitation peaks in the winter months, and becomes insignificant during the summer months.

------------------------------------------------------------------------

## Data Cleaning

As seen earlier, the data contains outliers and requires significant amounts of preprocessing.

This is a list of changes that were made to each dataset, and the reasons behind them:

### Stations

Dates were converted to POSIXct format \~ Enables easier manipulation and comparison later on

### Trips

Dates were converted to POSIXct format \~ Enables easier manipulation and comparison later on

Character fields were converted to factors \~ Enables grouping, and allows us to compare trends later on

Zip Code outliers were removed \~ Any value beyond the valid zip code range of 00501- 99950 was removed. Any blank values were converted to NAs, and the column was turned into integer data

Cancelled trips were identified \~ Any trip that lasted less than 3 minutes and started/ended at the same station was considered cancelled. These trips were removed from the data set and stored in a separate file. A snapshot of this file can be seen below (see the HTML version of this document to use an interactive version of the complete table):

```{r eval=TRUE, echo=FALSE, include=TRUE}
gt(head(read.csv("cancelled_trips.csv", header = T), 10)) %>% 
  tab_header(title = "Cancelled Trips")
```

Outliers were removed from the duration field \~ Any trip lasting longer than 1 day were removed from the data. It is assumed that any trip lasting longer than 1 day likely means that the customer forgot to return the bike.

### Weather

Date \~ Converted to POSIXct. This enables easier manipulation and comparison later on, as well as enables us to join to other data sets using date values as a common column

Events \~ Converted to factor

City \~ Converted to factor

Cloud_Cover\~ Converted to factor

Precipitation \~ Replaced trace values denoted by "T" with a small numeric value (0.005). Converted the column to be an integer field. This enables us to identify and compare precipitation trends.

------------------------------------------------------------------------

## Rush Hour Analysis

### Identifying Rush Hours:

To identify rush hour, the frequency of trips in each hour was plotted. In order to prevent biasing the "rush hour" to earlier or later times, the midpoint between the start and end time of each trip was calculated. This was chosen to give a representative estimate of peak hours. Further, rush hour can be defined as times of the day where trip frequency is above average. As such the mean frequency of trips was calculated (black line) and added to the plot for comparison's sake, as well as a means for objective identification of rush hours. Plotting the the frequency of trips by hour of the day shows us:

```{r plot_results, echo=FALSE, eval=TRUE, include=TRUE}
ggplot(weekdays, mapping = aes(x = midpoint)) +
  geom_histogram(fill = "lightseagreen", bins = 48) +
  scale_x_datetime(date_labels = "%H:%S",
                   breaks = "1 hour",
                   minor_breaks = "30 min") +
  theme(axis.text.x = element_text(size=7, angle = 70, vjust = 0.5),
        plot.title = element_text(hjust = 0.5)) +
  geom_hline(yintercept = meanfreq) +
  labs(title = "Trip Frequency by Hour",
       y = "Trips", x = "Time of Day")

```

Analyzing the plot, it is evident that the first set of rush hours are in the morning hours of 7am - 10:30pm. These hours see a large uptick in trip frequency, followed by the lunch rush from 11:30am - 1:30pm. The evening rush hour spans from about 6pm - 8pm.

### Busiest Stations During Rush Hours:

To identify the busiest stations during rush hours the data can be grouped by station, filtered by time (within rush hour boundaries established above), and the frequency of trips beginning at that station can be counted. These results can be visualized below:

```{r rush_hour_table, eval=TRUE, warning=FALSE, include=TRUE, paged.print=TRUE}

#determine 10 most frequent starting locations during rush hours
locations_wkdy %>%
  gt() %>% 
  tab_header(title = "Most Frequent Starting Locations",
             subtitle = "During Rush Hours")

#determine 10 most frequent ending locations during rush hours
locations_wkdy_end %>% 
  gt() %>% 
  tab_header(title = "Most Frequent Ending Stations",
             subtitle = "During Rush Hours")


```

The same methodology can be employed to identify the busiest stations on weekends:

```{r weekends, eval=TRUE, include=TRUE}

#determine 10 most frequent starting locations on weekends
locations_wknd %>%
  gt() %>% 
  tab_header(title = "Most Frequent Starting Stations",
             subtitle = "Weekends")

#determine 10 most frequent starting locations on weekends
locations_wknd_end %>%
  gt() %>% 
  tab_header(title = "Most Frequent Starting Stations",
             subtitle = "Weekends")

```

------------------------------------------------------------------------

### Utilization

In order to identify the degree of utilization of bikes, I decided to represent utilization as a percentage of the systems total utilization. This is calculated by identifying the total duration of rides in a month and expressing it as a proportion of the maximum duration the system can support. It should be noted that to calculate the total number of bikes in the system at a given time, the number of unique bike IDs in the trips data set were counted. This assumes that any bike that is not in use does not contribute to the total capacity of the system. It is likely that if a bike is not used at all in a given month, it is either broken, or being serviced. Another way to identify the number of bikes is to sum the number of docks at each station. The downside of this method is that it assumes that each station has as many bikes as there are docks, and that each of the bikes is used every month. For this reason I opted to calculate the value using the former method. The results of this analysis are visualized below:

```{r utilization, eval=TRUE, include=TRUE}
percent_utilization %>% 
  ggplot(mapping = aes(x = month, y= system_utilization, fill = system_utilization)) +
  geom_col() + 
  scale_fill_gradient(low = "red", high = "forestgreen", name = "System Utilization (%)")+
  labs(title = "Bike Rental System Utilization (%)",
       y = "",
       x = "Month")+
  theme(plot.title = element_text(hjust = 0.5),
        legend.position = "left",
        legend.title = element_text(hjust = 0.5, angle = 90),
        legend.title.position = "right")
```

From this figure it is evident that the system is more than capable of supporting the biking needs of it's population, as the maximum utilization is less than 2% of the total which occurs in the August. Consistent with the EDA data seen above, utilization peaks in the summer months and decreases in the winter months. To identify specifically how much each bike is used, I created an interactive table which summarizes each bike by the total duration of use (This can be accessed by cloning the repo and interacting with the table through the HTML file, otherwise the table below will just be a snapshot image). A specific ID can be searched and the usage per month can be identified. The table can also be sorted by pressing on the headers to identify bikes with the lowest usage that might be broken, lost or need servicing:

```{r bike_usage, eval=TRUE, include=T}

library(rmarkdown)

head(bike_data, 10) %>% 
  gt() %>% 
  tab_header(title = "Total Bike Usage")
```

------------------------------------------------------------------------

## Weather Analysis

To identify how the weather might impact ride patterns, data from the weather data set was joined to the trip data. For each city and date for which trip data exists, I joined with the weather conditions for that date/city. I then summarized each city/date with the total number of trips on that day as well as the total ride duration. I then created correlation matrices to identify how changes in weather conditions affected the total ride duration and total number of trips on a given day. These correlation plots identify the strength of linear correlation between 2 variables. A correlation may exist but this does not imply that it is statistically significant. For each correlation matrix, a p-value matrix was calculated to determine if a given correlation was statistically significant. Any results that were found to be insignificant (p \> 0.05 were removed from the plot. The results can be visualized for each city:\

```{r corplot, eval=TRUE,include=TRUE}
par(xpd=TRUE, mfrow = c(1,1))
for(i in unique(joined_data_numeric$city)){
  
  #create correlation matrix for the city
  cor_data <- joined_data_numeric %>%
    filter(city == i) %>% 
    select_if(is.numeric)
  
  #computes correlation values and p-values for the data
  cor_plot <- rcorr(as.matrix(cor_data))
  
  #clean the matrix to remove redundancy
  
  cor_plot$r <- cor_plot$r[1:2, c(-1,-2)]
  cor_plot$P <- cor_plot$P[1:2, c(-1,-2)]
  
  
  #plot the matrix, hide insignificant results (i.e H0: pearson coefficient = 0)
  corrplot(cor_plot$r, method = "shade", title = i, p.mat = cor_plot$P, sig.level = 0.05,insig = "blank",tl.col = "black", tl.srt = 50, cl.pos = "b", cl.ratio = 0.7, mar = c(2, 2, 2, 5))
}
```

The general trends that can be observed from the data are as follows:\
Total trip time and number of trips tend to correlate positively with better weather conditions. (i.e higher temperatures, better visibility and lower wind speeds). There seems to be a negative correlation between trip time, and trip frequency and precipitation. This is consistent with expectations as more rain likely discourages people from biking as opposed to other methods of transportation. It should be noted that the question marks on some of the plots represent correlations that could not be calculated. Notably the max visibility index often had a standard deviation of 0 preventing any meaningful correlation from being determined.
